---
title: "Biostat 203B Homework 1"
author: Yiyao Hu
subtitle: Due Jan 21 @ 11:59PM
output: html_document
---

```{r setup, include=FALSE}
  knitr::opts_chunk$set(echo = TRUE)
```

Display machine information for reproducibility:
```{r, eval=T}
  sessionInfo()
```

## Q1. Git/GitHub

**No handwritten homework reports are accepted for this course.** We work with Git and GitHub. Efficient and abundant use of Git, e.g., frequent and well-documented commits, is an important criterion for grading your homework.

1. Apply for the [Student Developer Pack](https://education.github.com/pack) at GitHub using your UCLA email. You'll get GitHub Pro account for free (unlimited public and private repositories).

2. Create a **private** repository `biostat-203b-2022-winter` and add `Hua-Zhou` and `maschepps` as your collaborators with write permission.

3. Top directories of the repository should be `hw1`, `hw2`, ... Maintain two branches `main` and `develop`. The `develop` branch will be your main playground, the place where you develop solution (code) to homework problems and write up report. The `main` branch will be your presentation area. Submit your homework files (R markdown file `Rmd`, `html` file converted from R markdown, all code and extra data sets to reproduce results) in `main` branch.

4. After each homework due date, teaching assistant and instructor will check out your main branch for grading. Tag each of your homework submissions with tag names `hw1`, `hw2`, ... Tagging time will be used as your submission time. That means if you tag your `hw1` submission after deadline, penalty points will be deducted for late submission.

5. After this course, you can make this repository public and use it to demonstrate your skill sets on job market.

## Q2. Data ethics training

This exercise (and later in this course) uses the [MIMIC-IV data](https://mimic-iv.mit.edu), a freely accessible critical care database developed by the MIT Lab for Computational Physiology. Follow the instructions at <https://mimic.mit.edu/docs/gettingstarted/> to (1) complete the CITI `Data or Specimens Only Research` course and (2) obtain the PhysioNet credential for using the MIMIC-IV data. Display the verification links to your completion report and completion certificate here. (Hint: The CITI training takes a couple hours and the PhysioNet credentialing takes a couple days; do not leave it to the last minute.)

**Solutions：** 

Link to my completion report: https://www.citiprogram.org/verify/?k469a8ce9-0f72-4fb6-8767-33c9eeaf0a9a-40438263

Link to my completion certificate:
Completion Certificate:
https://www.citiprogram.org/verify/?wcbf3112a-636f-46d1-85e8-31488931ef41-40438263


## Q3. Linux Shell Commands

1. The `/mnt/mimiciv/1.0` folder on teaching server contains data sets from MIMIC-IV. Refer to the documentation <https://mimic.mit.edu/docs/iv/> for details of data files.  
    ```{bash}
    ls -l /mnt/mimiciv/1.0
    ```
Please, do **not** put these data files into Git; they are big. Do **not** copy them into your directory. Do **not** decompress the gz data files. These create unnecessary big files on storage and are not big data friendly practices. Just read from the data folder `/mnt/mimiciv/1.0` directly in following exercises. 

    Use Bash commands to answer following questions.

2. Display the contents in the folders `core`, `hosp`, `icu`. Why are these data files distributed as `.csv.gz` files instead of `.csv` (comma separated values) files? Read the page <https://mimic.mit.edu/docs/iv/> to understand what's in each folder.

**Solution:** 

Why these data files are distributed as `.csv.gz` files instead of `.csv` is CSV is because, GZ is a CSV file compressed with gzip utility. 


The folder 'core' includes the information about "patient stay information (i.e. admissions and transfers)". Content of the folder 'core' is shown below:
    ````{bash}
    ls -l /mnt/mimiciv/1.0/core
    ````

The folder 'hosp' includes the information about "hospital level data for patients: labs, micro, and electronic medication administration". Content of the folder 'hosp' is shown below:
    ```{bash}
    ls -l /mnt/mimiciv/1.0/hosp
    ```
    

The folder 'icu' includes the information about "ICU level data. These are the event tables, and are identical in structure to MIMIC-III". Content of the folder 'icu' is shown below:
    ```{bash}
    ls -l /mnt/mimiciv/1.0/icu
    ```


    
3. Briefly describe what bash commands `zcat`, `zless`, `zmore`, and `zgrep` do.

**Solution:** 

View Compressed File and Uncompressed with zcat; 

Paging the compressed file with zless / zmore; 

Searching inside the compressed file with zgrep.



4. What's the output of following bash script?

**Solution:**

The output is shown below and it lists detailed contents of a directory.
    ```{bash, eval=T}
    for datafile in /mnt/mimiciv/1.0/core/*.gz
      do
        ls -l $datafile
      done
    ```

Display the number of lines in each data file using a similar loop.
    ```{bash, eval=T}
    for datafile in /mnt/mimiciv/1.0/core/*.gz
      do
        zcat $datafile | wc -l
      done      
    ```

5. Display the first few lines of `admissions.csv.gz`. How many rows are in this data file? How many unique patients (identified by `subject_id`) are in this data file? (Hint: combine Linux commands `zcat`, `head`/`tail`, `awk`, `sort`, `uniq`, `wc`, and so on.)

**Solution:**

The first five lines of `admissions.csv.gz`.
    ```{bash, eval=T}
    zcat  /mnt/mimiciv/1.0/core/admissions.csv.gz | head -5
    ```
  
    
Number of rows (without header) are in this data file：
    ```{bash, eval=T}
    zcat /mnt/mimiciv/1.0/core/admissions.csv.gz| tail -n +2| wc -l
    ```


Number of unique patients (without header):
    ```{bash, eval=T}
    zcat /mnt/mimiciv/1.0/core/admissions.csv.gz | awk -F, '{print $1}'| 
    tail -n +2| sort | uniq | wc -l
    ```


6. What are the possible values taken by each of the variable `admission_type`, `admission_location`, `insurance`, and `ethnicity`? Also report the count for each unique value of these variables. (Hint: combine Linux commands `zcat`, `head`/`tail`, `awk`, `uniq -c`, `wc`, and so on.)

**Solution:**

1) Possible values and corresponding counts of "admission_type" (field number=6)
    ```{bash, eval=T}
    zcat /mnt/mimiciv/1.0/core/admissions.csv.gz | awk -F, '{print $6}' |sort|
uniq -c
    ```


2) Possible values and corresponding counts of "admission_location" (field number=7)
    ```{bash, eval=T}
    zcat /mnt/mimiciv/1.0/core/admissions.csv.gz | awk -F, '{print $7}' |sort|
uniq -c
    ```
    

3) Possible values and corresponding counts of "insurance" (field number=9)
    ```{bash, eval=T}
    zcat /mnt/mimiciv/1.0/core/admissions.csv.gz | awk -F, '{print $9}' |sort|
uniq -c
    ```


4) Possible values and corresponding counts of "ethnicity" (field number=12)
    ```{bash, eval=T}
    zcat /mnt/mimiciv/1.0/core/admissions.csv.gz | awk -F, '{print $12}' |sort|
uniq -c
    ```



## Q4. Who's popular in Price and Prejudice

1. You and your friend just have finished reading *Pride and Prejudice* by Jane Austen. Among the four main characters in the book, Elizabeth, Jane, Lydia, and Darcy, your friend thinks that Darcy was the most mentioned. You, however, are certain it was Elizabeth. Obtain the full text of the novel from <http://www.gutenberg.org/cache/epub/42671/pg42671.txt> and save to your local folder. 
    ```{bash, eval=FALSE}
    wget -nc http://www.gutenberg.org/cache/epub/42671/pg42671.txt
    ```

Explain what `wget -nc` does. Do **not** put this text file `pg42671.txt` in Git. Complete the following loop to tabulate the number of times each of the four characters is mentioned using Linux commands.


'wget-nc' download pg42671.txt file into my HW1 folder from the URL address.  -nc means
--no-clobber
           If a file is downloaded more than once in the same directory,
           Wget's behavior depends on a few options, including -nc.  In
           certain cases, the local file will be clobbered, or overwritten,
           upon repeated download.  In other cases it will be preserved.

    ```{bash, eval=T}
    for char in Elizabeth Jane Lydia Darcy
    do
      echo $char:
      cat pg42671.txt|grep -n $char|wc -l
    done
    ```

2. What's the difference between the following two commands?
    ```{bash eval=FALSE}
    echo 'hello, world' > test1.txt
    ```
    and
    ```{bash eval=FALSE}
    echo 'hello, world' >> test2.txt
    ```
**Solution**:
'>' directs output from one commend to a file, and this comman will replace the previous contents in the file

'>>' appends output from one commend to a fill, it will add new content to the file under the orginal content. 

3. Using your favorite text editor (e.g., `vi`), type the following and save the file as `middle.sh`:
    ```{bash eval=FALSE}
    #!/bin/sh
    # Select lines from the middle of a file.
    # Usage: bash middle.sh filename end_line num_lines
    head -n "$2" "$1" | tail -n "$3"
    ```
Using `chmod` make the file executable by the owner, and run 
    ```{bash eval=FALSE}
    ./middle.sh pg42671.txt 20 5
    ```
Explain the output. Explain the meaning of `"$1"`, `"$2"`, and `"$3"` in this shell script. Why do we need the first line of the shell script? 

The #!/bin/sh is used to tell bash what shell you want to use to parse your command.
"$" - is just a sign of the shell prompt, means that shell is ready to accept commands. $1, $2, and $3 are the first three arguments of the shell script. 

## Q5. More fun with Linux

Try following commands in Bash and interpret the results: `cal`, `cal 2021`, `cal 9 1752` (anything unusual?), `date`, `hostname`, `arch`, `uname -a`, `uptime`, `who am i`, `who`, `w`, `id`, `last | head`, `echo {con,pre}{sent,fer}{s,ed}`, `time sleep 5`, `history | tail`.

**Solution:**

'Cal' shows the calender for current month

'cal 2021' shows the calender for all the 12 months in 2021

'cal 9 1752' shows the calender for September in 1975; Something unusual is that month lack the days from 9/13/1752 to 9/23/1752.

date: Look up today's date and time

hostname: It's used to obtain the DNS(Domain Name System) name and set the system's
hostname or NIS(Network Information System) domain name.

arch:It's used to print the computer architecture. 

uname -a:It's used to display basic information about the operating system and hardware. parameter a is used to reveal all the information

uptime: It is used to find out how long the system is active (running).

who am i:It displays the username of the current user when this command is invoked.

who: it's used to display the users currently logged in to your UNIX or Linux operating system.

w: It's used to show who is logged on and what they are doing. 

id: It's is used to find out user and group names and numeric ID's (UID or group ID) of the current user or any other user in the server. 

last | head:The last command in Linux is used to display the list of all the users logged in and out.The ahead command is used to display the first 10 results. 

echo {con,pre}{sent,fer}{s,ed}: It gives us the combination of all the possible words

time sleep 5 :Sleep command is used to delay for a fixed amount of time during the execution of any script.

history | tail: It is used to view the last 10th previously executed command.
    ```{bash eval=T}
     cal
     cal 2021
     cal 9 1752
    ```

    ```{bash}
    date
    ```

    ```{bash}
    hostname
    arch
    ```

    ```{bash}
    uname -a
    ```
    
    ```{bash}
   who am i
   who
   w
    ```

    ```{bash}
     id
    ```
    
    ```{bash}
    last | head
    ```
    
    ```{bash}
   echo {con,pre}{sent,fer}{s,ed}
    ```

    ```{bash}
    time sleep 5
    ```

    ```{bash}
    history |tail
    ```

